Dissertation Proposal

### 2023-11-07 proposal.tex
background: misconceptions

Students are given a working implementation of SMoL, built as a
\lstinline|#lang| language inside Racket
\citep{felleisenProgrammableProgrammingLanguage2018}. This language
provides only the defined syntax and semantics of SMoL, with no
(other) Racket features present. (As in Racket, arguments evaluate
left-to-right, to give stateful programs an unambiguous semantics.)
The implementation is available online:
\href{https://github.com/shriram/smol}{https://github.com/shriram/smol}.

@@

\begin{table}[t]
  \centering
  \begin{tabularx}{\textwidth}{>{\hsize=.6\hsize\RaggedRight}X >{\hsize=.8\hsize\RaggedRight}X >{\hsize=.5\hsize\RaggedRight}X >{\hsize=1.4\hsize\RaggedRight}X}
    \textbf{Publication}                                                           &
    \textbf{Population}                                                            &
    \textbf{Languages}                                                             &
    \textbf{Misconceptions}
    \\\hline
    \cite{fleuryParameterPassingRules1991}                                         &
    Unclear, likely CS2 students                                                   &
    Pascal                                                                         &
    \surmisedMiscon{CallerEnv}; \miscon{IsolatedFun}; \miscon{DefOrSet}
    (See their RULEs 2, 3a, and 3b in TABLE 2;
    RULE 1 doesn't apply to us.)
    \\\hline
    \cite{goldmanIdentifyingImportantDifficult2008,goldmanSettingScopeConcept2010} &
    CS1 students                                                                   &
    Java, Python, Scheme                                                           &
    Scope and memory model (See their Figures 4 and 5)
    \\\hline
    \cite{fislerAssessingTeachingScope2017}                                        &
    Third- and fourth-year undergrads                                              &
    Java and Scheme                                                                &
    \miscon{FlatEnv}; \miscon{CallByRef}; \miscon{CallsCopyStruct}; \miscon{DefByRef}
    (See their Section 4)
    \\\hline
    \cite{saarinenHarnessingWisdomClasses2019}                                     &
    CS2 students                                                                   &
    Java                                                                           &
    \miscon{StructByRef} (Their G2); \miscon{DefByRef} or \miscon{DefCopyStructs} (Their G3); \miscon{CallsCopyStruct} (Their G4).
    \\\hline
    \cite{strombackProgressionStudentsAbility2023}                                 &
    CS masters                                                                     &
    Python                                                                         &
    \miscon{FlatEnv}; \miscon{CallByRef}; \miscon{DefsCopyStruct}
    (See their section 4.2)
    \\\hline
    \cite{strombackProgressionStudentsAbility2023}                                 &
    CS undergrads                                                                  &
    C++                                                                            &
    \miscon{FlatEnv}; \miscon{CallByRef}; \miscon{DefsCopyStruct}
    (See their section 4.2)
    \\\hline
  \end{tabularx}
  \caption{Similar misconceptions found in prior research.}
  \label{t:rel-work}
\end{table}

Misconceptions related to scope, mutation, and higher-order functions have been
widely identified in varying populations
(from CS1 students to graduate students to users of online forums)
over many years (since at last 1991)
in varying programming languages (from Java to Racket) and in
different countries (such as the USA and Sweden).
\Cref{t:rel-work} lists works that seem most relevant to us.

There are some small differences.
\cite{fleuryParameterPassingRules1991} identifies a dynamic scope misconception
that is different from \miscon{FlatEnv}:
\begin{description}
  \item[\surmisedMiscon{CallerEnv}] Function values don't remember their environments.
    When a function is called, the function body is evaluated in an
    environment that extends from the \textit{caller's} environment.
\end{description}
I don't include \surmisedMiscon{CallerEnv} in our analysis because in our
data, all wrong answers that can be explained by \surmisedMiscon{CallerEnv}
are also explainable by \miscon{FlatEnv}.

Appendix A of \cite{sorvaVisualProgramSimulation2012} provides an
extensive survey of misconceptions reported
in research up to 2012.
There are overlaps between our survey and
theirs. For instance, our \surmisedMiscon{CallerEnv} is their No.~47.
However, because their descriptions are brief, it is difficult to tell
whether a misconception in their survey matches our misconceptions.
Because
they provide neither misinterpreters nor representative program-output
pairs, it is difficult to determine the overlaps precisely (showing
the value of providing these two machine-runnable descriptions).
At any rate,
we certainly find
no equivalent of \miscon{FlatEnv}, \miscon{DeepClosure}, and
\miscon{DefByRef} in their survey.

>>>>

Students are given a working implementation of SMoL, built as a
\lstinline|#lang| language inside Racket
\citep{felleisenProgrammableProgrammingLanguage2018}. This language
provides only the defined syntax and semantics of SMoL, with no
(other) Racket features present. (As in Racket, arguments evaluate
left-to-right, to give stateful programs an unambiguous semantics.)
The implementation is available online:
\begin{center}
  \href{https://github.com/shriram/smol}{https://github.com/shriram/smol}
\end{center}

@@

\begin{table}[t]
  \centering
  \begin{tabularx}{\textwidth}{>{\hsize=.6\hsize\RaggedRight}X >{\hsize=.8\hsize\RaggedRight}X >{\hsize=.6\hsize\RaggedRight}X >{\hsize=1.3\hsize\RaggedRight}X}
    \textbf{Publication}                                                           &
    \textbf{Population}                                                            &
    \textbf{Languages}                                                             &
    \textbf{Topics}
    \\\hline
    \cite{fleuryParameterPassingRules1991}                                         &
    Likely CS2 students                                                   &
    Pascal                                                                         &
    Scope
    \\\hline
    \cite{goldmanIdentifyingImportantDifficult2008,goldmanSettingScopeConcept2010} &
    CS1 students                                                                   &
    Java, Python, Scheme                                                           &
    Scope and memory model (See their Figures 4 and 5)
    \\\hline
    \cite{fislerAssessingTeachingScope2017}                                        &
    Third- and fourth-year undergrads                                              &
    Java and Scheme                                                                &
    Scope, variable aliasing, and structure aliasing.
    \\\hline
    \cite{saarinenHarnessingWisdomClasses2019}                                     &
    CS2 students                                                                   &
    Java                                                                           &
    Variable aliasing and structure aliasing.
    \\\hline
    \cite{strombackProgressionStudentsAbility2023}                                 &
    CS masters                                                                     &
    Python                                                                         &
    Scope, variable aliasing, and structure aliasing.
    \\\hline
    \cite{strombackProgressionStudentsAbility2023}                                 &
    CS undergrads                                                                  &
    C++                                                                            &
    Scope, variable aliasing, and structure aliasing.
    \\\hline
  \end{tabularx}
  \caption{Topics of misconceptions found in prior research.}
  \label{t:rel-work}
\end{table}

Misconceptions related to scope, mutation, and higher-order functions have been
widely identified in varying populations
(from CS1 students to graduate students to users of online forums)
over many years (since at last 1991)
in varying programming languages (from Java to Racket) and in
different countries (such as the USA and Sweden).
\Cref{t:rel-work} lists works that seem most relevant to us.
Appendix A of \cite{sorvaVisualProgramSimulation2012} provides an
extensive survey of misconceptions reported
in research up to 2012.

% There are some small differences.
% \cite{fleuryParameterPassingRules1991} identifies a dynamic scope misconception
% that is different from \miscon{FlatEnv}:
% \begin{description}
%   \item[\surmisedMiscon{CallerEnv}] Function values don't remember their environments.
%     When a function is called, the function body is evaluated in an
%     environment that extends from the \textit{caller's} environment.
% \end{description}
% I don't include \surmisedMiscon{CallerEnv} in our analysis because in our
% data, all wrong answers that can be explained by \surmisedMiscon{CallerEnv}
% are also explainable by \miscon{FlatEnv}.

% There are overlaps between our survey and
% theirs. For instance, our \surmisedMiscon{CallerEnv} is their No.~47.
% However, because their descriptions are brief, it is difficult to tell
% whether a misconception in their survey matches our misconceptions.
% Because
% they provide neither misinterpreters nor representative program-output
% pairs, it is difficult to determine the overlaps precisely (showing
% the value of providing these two machine-runnable descriptions).
% At any rate,
% we certainly find
% no equivalent of \miscon{FlatEnv}, \miscon{DeepClosure}, and
% \miscon{DefByRef} in their survey.
